---
title: "Getting Started with itsne"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with itsne}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5
)
```

The `itsne` package implements interval-valued extensions of t-SNE together with
baseline interval dimension-reduction methods, interval-aware LCMC evaluation
tools, and plotting helpers.

An *interval-valued* observation is a hyperrectangle rather than a point: each
variable is recorded as a `[lower, upper]` pair. Throughout the package this is
represented by paired columns ending in `_Lower` and `_Upper`.

This vignette uses small, fast examples. It is a demonstration of the package
interface, **not** a reproduction of the manuscript analyses.

## Bundled example data

The package ships the Face dataset and two interval-valued versions of the
Digits data in `inst/extdata`. Provenance, citations, licences, and checksums
for all three are recorded in `inst/extdata/README.md`.

```{r}
library(itsne)

face_path <- system.file("extdata", "facedata.csv", package = "itsne")
face_raw <- utils::read.csv(face_path, check.names = FALSE)
face_data <- face_raw[, c("species", grep("_(Lower|Upper)$", names(face_raw), value = TRUE))]
face_std <- standardize_interval_data(face_data, id_col = "species", method = 1)
head(face_std[, 1:5])
```

## Baseline interval embeddings

The baseline interval PCA methods are inexpensive and well suited to a first
look.

```{r}
ipca_fit <- ipca_qm(face_std, id_col = "species", m = 5, dims = 2)
head(ipca_fit)
```

```{r}
plot_interval_projection(ipca_fit, title = "Face data: IPCA(QM)")
```

## The four I-tSNE variants

Two variants are *expansion-based*: each interval is replaced by representative
points, standard t-SNE is applied to the pooled cloud, and a reduced-space
interval is reconstructed afterwards as a bounding box.

```{r}
sim_data <- simulate_point_aggregation(
  cluster_centers = matrix(c(6, 0, -6, 0, 0, 6), ncol = 2, byrow = TRUE),
  n_per_cluster = 30,
  k = 2,
  noise_dims = 0,
  seed = 1
)

qm_fit <- itsne_qm(sim_data, id_col = "Group", m = 3, perplexity = 5,
                   max_iter = 250)
head(qm_fit)
```

Two variants are *direct*: they optimize the reduced-space interval itself.
`itsne_cr()` uses a center-radius parameterization that guarantees positive
radii, so every returned interval is valid by construction.

```{r}
cr_fit <- itsne_cr(sim_data, id_col = "Group", perplexity = 5,
                   max_iter = 250, verbose = FALSE)
head(cr_fit$embedding)
```

### Endpoint order in `itsne_mm()`

`itsne_mm()` optimizes coupled lower and upper endpoints under a **soft** order
penalty. For any finite `penalty_lambda` this does not guarantee
`lower <= upper` at a local numerical solution, so the fit must be checked
afterwards. The package reports violations but **never repairs them**: no
reordering, swapping, sorting, clipping, or coordinatewise hull is applied, and
`embedding` always holds the raw optimizer output.

```{r}
mm_fit <- itsne_mm(sim_data, id_col = "Group", perplexity = 5,
                   max_iter = 250, penalty_lambda = 0.2, verbose = FALSE)

mm_fit$violation_summary$n_violations
mm_fit$violation_summary$max_violation
mm_fit$violation_summary$repaired
```

When at least one coordinate is inverted, a warning is emitted as well. If that
happens, raise `penalty_lambda` and refit, or use `itsne_cr()` instead.

## Interval-aware LCMC evaluation

Neighborhood preservation is assessed with the Local Continuity Meta-Criterion,
computed under interval-valued distances in both the original and reduced
spaces.

```{r}
methods <- list(
  ipca_qm = ipca_qm(sim_data, id_col = "Group", m = 3, dims = 2),
  ipca_cr = ipca_cr(sim_data, id_col = "Group", dims = 2)
)

lcmc_tables <- compute_lcmc_tables(sim_data, methods, k_range = 1:3)
plot_lcmc_grid(lcmc_tables)
```

## End-to-end workflows

Two convenience entry points run larger analysis sequences and write figures and
summary tables to disk. They are considerably more expensive than the examples
above and are not run when this vignette is built.

By default they write to a subdirectory of `tempdir()`; pass `output_dir` to
choose a location. They never write into the working directory unless you ask
them to.

```{r, eval = FALSE}
out <- run_simulation_studies(output_dir = file.path(tempdir(), "itsne-sim"))
run_real_data_analysis(output_dir = file.path(tempdir(), "itsne-real"))
```

These are intended for reproducibility runs; the lower-level modeling and
plotting functions are usually the better interface for exploratory work.
