%\VignetteIndexEntry{Getting Started with itsne}
%\VignetteEngine{knitr::knitr}
%\VignetteEncoding{UTF-8}

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5
)
```

# Getting Started with `itsne`

The `itsne` package implements interval-valued extensions of t-SNE together with baseline interval dimension-reduction methods, modified-LCMC evaluation tools, and plotting helpers used in the accompanying manuscript.

## Bundled example data

The package ships with the Face dataset and two interval-valued versions of the Digits analysis data in `inst/extdata`. The example below reads the Face data and standardizes the interval variables.

```{r}
library(itsne)
library(readr)

face_path <- system.file("extdata", "facedata.csv", package = "itsne")
face_raw <- read_csv(face_path, show_col_types = FALSE)
face_data <- face_raw[, c("species", grep("_(Lower|Upper)$", names(face_raw), value = TRUE))]
face_std <- standardize_interval_data(face_data, id_col = "species", method = 1)
head(face_std[, 1:5])
```

## Baseline interval embeddings

For a quick reduced-space representation, the baseline IPCA methods are inexpensive and well suited for exploratory use.

```{r}
ipca_fit <- ipca_qm(face_std, id_col = "species", m = 5, dims = 2)
head(ipca_fit)
```

```{r}
plot_interval_projection(ipca_fit, title = "Face data: IPCA(QM)")
```

## Modified-LCMC evaluation

The package also computes the modified Local Continuity Meta-Criterion (LCMC) used throughout the manuscript.

```{r}
sim_data <- simulate_point_aggregation(
  cluster_centers = matrix(c(6, 0, -6, 0, 0, 6), ncol = 2, byrow = TRUE),
  n_per_cluster = 30,
  k = 2,
  noise_dims = 0,
  seed = 1
)

methods <- list(
  ipca_qm = ipca_qm(sim_data, id_col = "Group", m = 3, dims = 2),
  ipca_cr = ipca_cr(sim_data, id_col = "Group", dims = 2)
)

lcmc_tables <- compute_lcmc_tables(sim_data, methods, k_range = 1:3)
plot_lcmc_grid(lcmc_tables)
```

## I-tSNE methods and end-to-end workflows

The direct I-tSNE methods are more computationally intensive than the baseline examples shown above, so they are often better run as part of a dedicated analysis script.

```{r, eval = FALSE}
itsne_fit <- itsne_qm(face_std, id_col = "species", m = 5, perplexity = 20)
plot_interval_projection(itsne_fit, title = "Face data: I-tSNE(QM)")
```

The package includes two convenience entry points that reproduce the manuscript workflows and write figures and summary tables to disk.

```{r, eval = FALSE}
run_simulation_studies(output_dir = "results/simulations")
run_real_data_analysis(output_dir = "results/real-data")
```

These functions are intended for reproducibility runs, while the lower-level modeling and plotting functions are usually the best interface for exploratory analysis and method development.